---
title: "Tidy Data and Visualisation with R Exercises"
mainfont: Ubuntu
sansfont: Ubuntu
monofont: Courier New
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
  word_document: default
editor_options:
  chunk_output_type: console
params:
  include_output: FALSE
---

```{r load_libraries, include=FALSE, message=FALSE}
# load packages
packages <-  c("knitr", "rprojroot")
for( package in packages ){
  library( package, character.only = TRUE )
}
```

```{r knitr_options, include=FALSE, message=FALSE}
opts_chunk$set( fig.width=9, fig.height=6, include=params$include_output, echo=params$include_output, 
                results = params$include_output, warning = FALSE, message = FALSE,
                fig.show = params$include_output)

# create logical variable for whether to include page breaks
# TRUE when output is latex AND include_output is TRUE
include_breaks <- opts_knit$get('rmarkdown.pandoc.to') == 'latex' &
  params$include_output
```

Before you start, make sure you’ve read the document that describes the zebrafish dataset we’re using in this practical. And make sure you’ve put the required file (`Amp.counts.tsv`) in your home directory.

Load the example dataset (`Amp.counts.tsv`) using `read_tsv`.

```{r tidyverse, include = TRUE, echo = TRUE, eval=FALSE}
library(tidyverse)
results <- read_tsv('Amp.counts.tsv',
                    col_types = cols(Chr = col_character(), 
                                     Strand = col_character()))
```

```{r load_data, echo = FALSE, eval=TRUE}
library(tidyverse)
results <- read_tsv('data/Amp.counts.tsv',
                    col_types = cols(Chr = col_character(), 
                                     Strand = col_character()))
```

## Tidy Data

1.  Use `filter` to find out how many genes have an adjusted p-value less than 0.05.

```{r filtering}
filter(results, adjp < 0.05) %>%
  nrow()
```

2.  Find out which gene has the smallest p-value by sorting the data using `arrange`.

```{r arrange}
arrange(results, adjp) %>%
  select(., Gene, Name, adjp) %>%
  head(1)
```

3.  Make a new column in the data that is -log10 of the adjusted p-value column. You can use the `log10()` function to calculate this.

```{r mutate}
results <- mutate(results, 
       log10p = -log10(adjp))

select(results, Gene, Name, adjp, log10p) %>% 
  arrange(adjp) %>% 
  head(5)
```

```{r results='asis', eval=include_breaks, echo = FALSE}
cat('\\pagebreak')
```

4.  Make a new data.frame that contains the Gene, Name and all the normalised count columns.

```{r select}
normalised_counts <- 
  select(results, Gene, Name, contains('normalised count'))
```

5.  Make the new data.frame data tidy using `pivot_longer`.

```{r pivot}
normalised_counts_long <- 
  pivot_longer(normalised_counts, c(-Gene, -Name),
               names_to = 'sample', 
               values_to = 'normalised count') %>% 
  # this removes the " normalised count" suffix from
  # each count column name to leave just the sample name
  mutate(sample = sub(" normalised count", "", sample))
```

```{r results='asis', eval=include_breaks, echo = FALSE}
cat('\\pagebreak')
```

## Plotting

Load the penguins dataset with `library(palmerpenguins)`.

1.  Using the `penguins` dataset, make a plot of body mass (body_mass_g) against bill length (bill_length_mm).

```{r plot1}
library(palmerpenguins)
ggplot(data = penguins) + 
  geom_point(aes(x = body_mass_g, y = bill_length_mm))
```

```{r results='asis', eval=include_breaks, echo = FALSE}
cat('\\pagebreak')
```

2.  Now colour the points by flipper length (flipper_length_mm) and use the `viridis` colour scale using `scale_colour_viridis_c`.

```{r plot2}
ggplot(data = penguins) + 
  geom_point(aes(x = body_mass_g, y = bill_length_mm,
                 colour = flipper_length_mm)) +
  scale_colour_viridis_c()
```

```{r results='asis', eval=include_breaks, echo = FALSE}
cat('\\pagebreak')
```

3.  Change the shape of the points to a hollow shape (one of 21-25).  
Make species the fill colour, and pick 3 colours to use with `scale_fill_manual`.

```{r plot3}
colour_blind_palette <- 
  c("#0073B3", "#CC6600", "#009980", "#F2E640", 
    "#59B3E6", "#CC99B3", "#000000", "#E69900")

ggplot(data = penguins) + 
  geom_point(aes(x = body_mass_g, y = bill_length_mm,
                 fill = species),
             shape = 22, size = 3) +
  scale_fill_manual(values = colour_blind_palette)

```

```{r results='asis', eval=include_breaks, echo = FALSE}
cat('\\pagebreak')
```

`r ifelse(params$include_output, "Alternatively you could specify the fill colours as names", "")`

```{r plot3_names}
ggplot(data = penguins) + 
  geom_point(aes(x = body_mass_g, y = bill_length_mm,
                 fill = species),
             shape = 22, size = 3) +
  scale_fill_manual(values = c('firebrick2', 'steelblue3', 
                               'goldenrod1'))
```
